---
title: "Plotly Exploration"
author: "Eric Erkela"
date: "11/24/2020"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


## THOR Data
Theater History of Operations (THOR), is a painstakingly cultivated database of historic aerial bombings from World War I through Vietnam. The value of THOR is immense, and has already proven useful in finding unexploded ordinance in Southeast Asia and improving Air Force combat tactics. Our goal is to see where public discourse and innovation takes this data.

This dataset combines digitized paper mission reports from WWII. It can be searched by date, conflict, geographic location and more than 60 other data elements to form a live-action sequence of the air war from 1939 to 1945. The records include U.S. and Royal Air Force data, as well as some Australian, New Zealand and South African air force missions.

For more information, see: https://data.world/datamil/world-war-ii-thor-data

``` {r, echo = FALSE, message = FALSE, warning = FALSE}
library(dplyr, quietly = TRUE, warn.conflicts = FALSE)
library(tidyr, quietly = TRUE, warn.conflicts = FALSE)
library(plotly, quietly = TRUE, warn.conflicts = FALSE)
library(zoo, quietly = TRUE, warn.conflicts = FALSE)
```

***
``` {r, echo = FALSE, warning = FALSE}
thor_data <- read.csv("THOR_WWII_DATA_CLEAN.csv")
thor_data$MSNDATE <- as.Date(thor_data$MSNDATE, format = "%m/%d/%Y")

country_names <- c("ALBANIA", "ALEUTIAN ISLANDS", "ALGERIA", "ANDAMAN ISLANDS",
                   "AUSTRALIA", "AUSTRIA", "BALI", "BELGIUM",
                   "BISMARK ARCHIPELAGO", "BORNEO", "BOUGAINVILLE", "BULGARIA",
                   "BURMA", "CAROLINE ISLANDS", "CELEBES ISLANDS", "CHINA",
                   "CHINA MINING", "CORAL SEA AREA", "CORSICA", "CRETE",
                   "CYPRUS", "CZECHOSLOVAKIA", "DENMARK", "EGYPT", "ERITREA",
                   "ETHIOPIA", "ETHIOPIA/ABSINNYA", "FORMOSA",
                   "FORMOSA AND RYUKYU ISLANDS", "FRANCE", "FRENCH INDO CHINA",
                   "FRENCH INDO CHINA MINING", "FRENCH WEST AFRICA", "GERMANY",
                   "GILBERT ISLANDS", "GREAT BRITAIN", "GREECE",
                   "HOLLAND OR NETHERLANDS", "HUNGARY", "INDIA", "INDIAN OCEAN",
                   "INDONESIA", "IRAQ", "ITALY", "JAPAN", "JAPAN MINING",
                   "JAVA", "KOREA OR CHOSEN", "KOREA OR CHOSEN MINING",
                   "KURILE ISLANDS", "LEBANON", "LIBYA", "LUXEMBOURG",
                   "MADAGASCAR", "MALAY STATES", "MALAY STATES MINING",
                   "MANCHURIA", "MARCUS ISLANDS", "MARIANAS ISLANDS",
                   "MARSHALL ISLANDS", "MOROCCO", "NETHERLANDS EAST INDIES",
                   "NEW GUINEA", "NEW IRELAND", "NORWAY", "PALAU ISLANDS",
                   "PANTELLARIA", "PHILIPPINE ISLANDS", "POLAND", "ROMANIA",
                   "SARDINIA", "SICILY", "SOLOMON ISLANDS", "SOMALIA", "SUDAN",
                   "SUMATRA", "SUMATRA MINING", "SWITZERLAND", "SYRIA",
                   "THAILAND OR SIAM", "THAILAND OR SIAM MINING", "TIMOR",
                   "TUNISIA", "TURKEY", "VOLCANO AND BONIN ISLANDS",
                   "WAKE ISLAND", "YUGOSLAVIA")

country_codes <- c("AFG", "USA", "DZA", "IND", "AUS", "AUT", "IDN", "BEL",
                   "PNG", "IDN", "PNG", "BGR", "MMR", "FSM", "IDN", "CHN",
                   "CHN", "AUS", "FRA", "GRC", "CYP", "CZE", "DNK", "EGY",
                   "ERI", "ETH", "ETH", "TWN", "TWN", "FRA", "VNM", "VNM",
                   "CIV", "DEU", "FSM", "GBR", "GRC", "NLD", "HUN", "IND",
                   "IOT", "IDN", "IRQ", "ITA", "JPN", "JPN", "IDN", "KOR",
                   "KOR", "RUS", "LBN", "LBY", "LUX", "MDG", "MYS", "MYS",
                   "CHN", "JPN", "GUM", "MHL", "MAR", "IDN", "PNG", "PNG",
                   "NOR", "PLW", "ITA", "PHL", "POL", "ROU", "ITA", "ITA",
                   "SLB", "SOM", "SDN", "IDN", "IDN", "CHE", "SYR", "THA",
                   "THA", "TLS", "TUN", "TUR", "JPN", "USA", "SRB")

country_conv <- data.frame(name = country_names,
                           code = country_codes)
thor_data$TGT_COUNTRY_CODE <- as.factor(unlist(sapply(thor_data$TGT_COUNTRY,
                                     function(i) {
                                       if (i %in% country_conv$name) {
                                         country_conv[country_conv$name == i, 2]
                                       } else {
                                         NA
                                       }
                                     })))

raids_by_country <- thor_data %>%
  mutate(INDEX = 1) %>%
  group_by(TGT_COUNTRY_CODE) %>%
  summarize(NUM_RAIDS = sum(INDEX), 
            ORDNANCE = sum(TOTAL_TONS, na.rm = TRUE),
            .groups = "drop") %>%
  mutate(TEXT = paste0("Total Ordnance Dropped: <br>", 
                       ORDNANCE,
                       " tons")) %>%
  arrange(desc(NUM_RAIDS))

g1 <- plot_ly(raids_by_country, 
              type = "choropleth", 
              locations = raids_by_country$TGT_COUNTRY_CODE,
              z = log10(raids_by_country$ORDNANCE + 1),
              text = raids_by_country$TEXT,
              colorscale = "reds",
              marker = list(line = list(color = toRGB("gray30"))),
              hoverinfo = "name+location+text",
              showlegend = FALSE,
              showscale = FALSE) %>%
  layout(title = "Allied Ordnance Dropped on Each Target Country (WWII)")
g1
```

***
``` {r, echo = FALSE, warning = FALSE}
major_theaters <- thor_data %>%
  group_by(TGT_COUNTRY) %>%
  summarize(ORDNANCE = sum(TOTAL_TONS, na.rm = TRUE),
            .groups = "drop") %>%
  arrange(desc(ORDNANCE))
major_theaters <- major_theaters$TGT_COUNTRY[1:10]
  
ordnance_by_country <- thor_data %>%
  mutate(MSNDATE = as.yearmon(MSNDATE)) %>%
  group_by(MSNDATE, TGT_COUNTRY) %>%
  summarize(ORDNANCE = sum(TOTAL_TONS, na.rm = TRUE),
            .groups = "drop") %>%
  filter(TGT_COUNTRY %in% major_theaters)

g2 <- ggplot(data = ordnance_by_country, aes(x = MSNDATE, y = ORDNANCE)) +
  geom_histogram(stat = "identity", aes(fill = TGT_COUNTRY)) + 
  labs(title = "Ordnance Dropped in Allied Bombing Raids by Target Country (WWII)",
       x = "Date (Month/Year)", y = "Ordnance Dropped (tons)")
ggplotly(g2)
```