---
title: "Statistical Inference Project Part 1: Simulation Exercise"
author: "Eric Erkela"
date: "11/16/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Overview
In this section of the two-part statistical inference project, we will investigate the Central Limit Theorem (CLT) as it relates to estimates of population means taken from a simulated data set.  From a randomly-generated, exponential data set, we will show that the distribution of sample averages is approximately normal, with mean and variance identical to the exponential distribution from which they were taken. 

## Dependencies
Below are all the packages used in this analysis, produced here for convenience:

``` {r, warning = FALSE}
library(dplyr, warn.conflicts = FALSE)
library(ggplot2)
```

# Part 1: Simulations
Before we start analyzing, we need to simulate our raw data.  This will consist of 1000 sample means, each taken from a set of 40 random observations of an exponential distribution (lambda = 0.2).

``` {r}
set.seed(123)       # for reproducibility
n <- 1000; b <- 40  # number of averages, number of observations per average
lambda <- 0.2       # exponential rate, determined in project instructions
sim_data <- matrix(rexp(n * b, rate = lambda), nrow = n, ncol = b)
sim_means <- apply(sim_data, 1, mean)
head(sim_means)
```

Since our data comes from an exponential distribution for which we know lambda, we can exactly calculate its expected mean/variance.  How do these compare to what we've observed?

``` {r}
data.frame(Mean = c(mean(sim_means), 1 / lambda),
           Variance = c(var(sim_means), (1 / lambda)^2 / b),
           row.names = c("Observed", "Expected"))
```

Pretty close! This shows that our simulated data does in fact estimate the theoretical population mean/variance of the exponential distribution it is based on, a basic tenet of statistical inference.  If we were to plot the probability density of our sample means by their value, what distribution would we get?  A naïve guess might be that they would match the distribution from which they were taken.  In this scenario, our data would look something like this, where we've substituted our sample means for the first column of our simulated data set.

``` {r}
g <- ggplot(data = tibble(value = sim_data[, 1]), aes(x = value))
g + 
  geom_histogram(aes(y = ..density..),
                 bins = 20) + 
  geom_vline(xintercept = mean(sim_data[, 1]),
             linetype = "dashed",
             color = "black",
             size = 1) +
  stat_function(fun = dexp, 
                n = 101,
                args = list(rate = lambda),
                color = "red",
                size = 1) + 
  labs(title = "Simulated Exponential Distribution")
```

Makes sense, right?  But what do we actually get?

``` {r}
g <- tibble(value = sim_means) %>% ggplot(., aes(x = value))
g + 
  geom_histogram(aes(y = ..density..),
                 bins = 20) + 
  geom_vline(xintercept = mean(sim_means),
             linetype = "dashed",
             color = "black",
             size = 1) + 
  stat_function(fun = dexp, 
                n = 101,
                args = list(rate = lambda),
                color = "red",
                size = 1) + 
  stat_function(fun = dnorm, 
                n = 101,
                args = list(mean = mean(sim_means), sd = sd(sim_means)),
                color = "blue",
                size = 1) + 
  labs(title = "Density of Simulated Means from Exponential Distribution (lambda=0.2)",
       x = "Simulated Mean",
       y = "Density")
```

As we can see, our sample means are clearly not exponentially distributed.  Instead, we find that they follow a normal distribution (blue line) with mu = mean(sim_means) and sd = sd(sim_means).  This matches what we'd expect from the CLT.