---
title: "Statistical Inference Project (Part 1)"
author: "Eric Erkela"
date: "11/16/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Overview
In this section of the two-part statistical inference project, we will investigate the Central Limit Theorem (CLT) as it relates to estimates of population means taken from a simulated data set.  From a randomly-generated, exponential data set, we will show that the distribution of sample averages is approximately normal, with mean and variance identical to the exponential distribution from which they were taken. 

## Dependencies
Below are all the packages used in this analysis, produced here for convenience:

``` {r, warning = FALSE}
library(dplyr, warn.conflicts = FALSE)
library(ggplot2)
```

# Part 1: Simulations
In the first part of this assignment, we will be applying the Central Limit Theorem (CLT) to show the Gaussian convergence of a set of sample means, each taken from 40 random observations of an exponential distribution(lambda = 0.2).  Before we do, however, we must generate the data itself:

``` {r}
set.seed(123)       # for reproducibility
n <- 1000; b <- 40  # number of averages, number of observations per average
lambda <- 0.2       # exponential rate, determined in project instructions
sim_data <- matrix(rexp(n * b, rate = lambda), nrow = n, ncol = b)
sim_means <- apply(sim_data, 1, mean)
head(sim_means)
```

Et voilà! Since our data comes from an exponential distribution for which we know lambda, we can exactly calculate its expected mean/variance.  How do these compare to what we've observed?

``` {r}
data.frame(Mean = c(mean(sim_means), 1 / lambda),
           Variance = c(var(sim_means), (1 / lambda)^2 / b),
           row.names = c("Observed", "Expected"))
```

Pretty close! This shows that our simulated data does in fact estimate the theoretical population mean/variance of the exponential distribution it is based on, a basic tenet of statistical inference.  If we were to plot the probability density of our sample means by their value, what distribution would we get?  A naïve guess might be that they match the distribution from which they were taken, an exponential with lambda = 0.2.  In this scenario, our data would look something like this, where we've substituted our sample means for the first column of our original, simulated data set.

``` {r}
g <- ggplot(data = tibble(value = sim_data[, 1]), aes(x = value))
g + 
  geom_histogram(aes(y = ..density..),
                 bins = 20) + 
  geom_vline(xintercept = mean(sim_data[, 1]),
             linetype = "dashed",
             color = "black",
             size = 1) +
  stat_function(fun = dexp, 
                n = 101,
                args = list(rate = lambda),
                color = "red",
                size = 1)
```

Makes sense, right?  But what do we actually get?

``` {r}
g <- tibble(value = sim_means) %>% ggplot(., aes(x = value))
g + 
  geom_histogram(aes(y = ..density..),
                 bins = 20) + 
  geom_vline(xintercept = mean(sim_means),
             linetype = "dashed",
             color = "black",
             size = 1) + 
  stat_function(fun = dexp, 
                n = 101,
                args = list(rate = lambda),
                color = "red",
                size = 1) + 
  stat_function(fun = dnorm, 
                n = 101,
                args = list(mean = mean(sim_means), sd = sd(sim_means)),
                color = "blue",
                size = 1) + 
  labs(title = "Density of Simulated Means from Exponential Distribution (lambda=0.2)",
       x = "Simulated Mean",
       y = "Density")
```

As we can see, our sample means are clearly not exponentially distributed (the red line).  Instead, we find that they follow a normal distribution (blue line) with mu = mean(sim_means), which matches the dashed black line in the figure above, and sd = sd(sim_means).  This matches what we'd expect from the CLT.